Race to the Ballot Box: Will Demographics Decide a Winner This Year?

### 2024-11-02 paper/paper.qmd
Updating the model, results, and appendix

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Ut elit leo, viverra ac enim non, pulvinar aliquam nisl. Nam accumsan ac nisi at consectetur. Curabitur luctus lacus eget risus rutrum, vel facilisis urna feugiat. Ut ac dictum velit, in blandit augue. Fusce sit amet vehicula dui. Suspendisse eleifend tempor rhoncus. Cras vehicula, nisl et molestie pulvinar, lorem ex dictum augue, nec mollis leo sem quis mauris. In lacus massa, maximus vel vehicula in, tincidunt sed lorem. Proin commodo pulvinar dictum. Vestibulum eleifend, metus ac luctus molestie, velit odio cursus nisl, sed lobortis arcu massa id diam. Curabitur convallis erat a ultricies dictum. In imperdiet ultricies nibh, quis condimentum nulla mattis non.

Aliquam erat volutpat. Praesent in erat sit amet purus auctor posuere ut vitae leo. Suspendisse potenti. Interdum et malesuada fames ac ante ipsum primis in faucibus. Ut blandit vel orci vel aliquam. Curabitur a turpis vestibulum enim interdum consectetur at et dui. Praesent ut mauris maximus, auctor nisl ac, eleifend sapien. Morbi non varius urna. Ut placerat vestibulum nisi in lacinia. Aliquam non ultrices arcu, et rhoncus dolor. Suspendisse vel pulvinar tortor. Nulla lacinia sapien sollicitudin, vestibulum nulla id, hendrerit nisl.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#| label: tbl-prediction
#| echo: false
#| 
library(tidyverse)
library(rstanarm)
library(caret)
library(kableExtra)
library(arrow)

@@

predicted_data_harris |> 
  merge(predicted_data_trump, all = T) |>
  left_join(electoral_votes, by = c("state" = "state")) |>
  mutate(predicted_winner = ifelse(replace_na(predicted_pct_harris, 50) <
                                     replace_na(predicted_pct_trump, 50), 
                                   "Trump", "Harris")) |>
  kbl(booktabs = TRUE, col.names = c("State", "Predicted Harris %", "Predicted Trump %", "Electoral Votes", "Predicted Winner"))
```

@@

@@

# Appendix B: Idealized Methodology and Survey

>>>>

The table below summarizes the predicted support percentages for Donald Trump and Kamala Harris across the states where polling data is available. Due to the model being trained on a subset of the cleaned data and the specific approach used to split training and testing data, there are some missing values for each candidate. These missing values are expected and are discussed in further detail in the appendix. Additionally, in some cases, the sum of the predicted percentages for Trump and Harris slightly exceeds 100%, falling within a typical margin of error. This minor discrepancy is likely due to prediction error, which is also explored in the appendix.

Based on the predicted percentages, a winning candidate is identified for each state, assuming the winner takes all electoral votes for that state. As anticipated, both candidates are expected to secure wins in their respective stronghold states, with no predicted changes in non-swing states. This highlights the critical role of swing states, where a victory for either candidate could be decisive in the overall election outcome. The results underline the importance for both Trump and Harris to perform well in these battleground regions to secure their path to victory.

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align='center'}
#| label: tbl-prediction
#| echo: false
#| 
library(tidyverse)
library(rstanarm)
library(caret)
library(kableExtra)
library(arrow)

@@

predicted_data_harris |> 
  merge(predicted_data_trump, all = T) |>
  left_join(electoral_votes, by = c("state" = "state")) |>
  mutate(predicted_winner = ifelse(replace_na(predicted_pct_harris, 50) <
                                     replace_na(predicted_pct_trump, 50), 
                                   "Trump", "Harris")) |>
  kbl(booktabs = TRUE, col.names = c("State", "Harris %", "Trump %", "Electoral Votes", "Winner"))
```

@@

The model evaluation relies on the F1 Score and RMSE (Root Mean Squared Error) as these metrics capture key aspects of prediction accuracy and reliability, aligning well with the paper’s objectives. The F1 Score, as the harmonic mean of precision and recall, is particularly useful here because it balances the model’s ability to correctly identify strong support for either candidate, ensuring both Trump and Harris’s predicted support is captured with minimal false positives and false negatives. This is essential in a political forecasting model, where the correct classification of state support is critical for predicting election outcomes. Meanwhile, RMSE provides insight into the overall prediction accuracy by measuring the average deviation of predictions from actual values. This helps gauge how well the model can approximate polling percentages, especially for swing states where slight errors can significantly impact projected outcomes. Together, these metrics ensure that the model is both precise in classification and accurate in magnitude, supporting the paper’s goal of making robust, actionable election predictions.

# Additional Results Details

## Null Values

Some states in the summary table have null values in the predicted percentages for each candidate, which is due to the subsampling and splitting approach used in the model training process. Specifically, the model was trained on a subset of the cleaned dataset found in `data/analysis_data/analysis_data.parquet`, and when training and testing data were split, certain states did not have sufficient representation in both subsets. This led to gaps in the predictions for some states, as the model could not generalize well for locations with limited or no data during training. While these null values do not affect the core results, they highlight a limitation in the sampling approach, especially in states with fewer polling data points, and are further discussed with suggestions for improvement in the appendix.

## Prediction Errors

In some cases, the predicted percentages for Trump and Harris sum to slightly more than 100%, which reflects prediction error within an acceptable margin. This overestimation is consistent with the margin of error typically seen in polling averages, indicating that while the model is effective overall, minor discrepancies arise from rounding and the inherent variability in polling data. These errors can also result from the weights applied to pollsters based on their reliability scores, which may amplify certain pollster biases. To maintain accuracy, the model incorporates error margins similar to those in the polling data, ensuring predictions are within a reasonable range, with further analysis and adjustments explained in the appendix.

@@

# Idealized Methodology and Survey
